ranacapa: An R package and Shiny web app to explore environmental DNA data with exploratory statistics and interactive visualizations

### 2018-10-14 manuscript.Rmd
latest tweaks

---
title: 'ranacapa: An R package to explore environmental DNA data with exploratory statistics and interactive visualizations'
author: Gaurav S. Kandlikar (1,+), Zachary J. Gold (1), Madeline C. Cowen (1), Rachel  S. Meyer (1), Amanda C. Freise (2), Nathan J.B. Kraft (1), Jordan Moberg-Parker (2), Joshua Sprague (3), David Kushner (3), and Emily E. Curd (1)
fontsize: 11pt
output: 
  word_document:
    reference_docx: rendering-templates/word_template.docx
urlcolor: blue
always_allow_html: yes
bibliography: bibliography.bib
csl: rendering-templates/ecology-letters.csl
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
load("make-figures.Rdata")
options(digits = 3)
```

@@

### Affiliations and Contact Information  
(1) Department of Ecology and Evolutionary Biology, University of California – Los Angeles   
(2) Department of Microbiology and Molecular Genetics, University of California – Los Angeles  
(3) National Park Service

@@

**Keywords**: environmental DNA; data visualization; citizen science; community science; shiny; metabarcoding; education; community ecology

@@

## Abstract 
```{r child = 'abstract.md'}
```

@@

The targeted amplification and sequencing of DNA that living organisms shed into their physical environment, termed "environmental DNA (eDNA) metabarcoding",  is revolutionizing microbiology, ecology, and conservation research [@Taberlet_2012; @Deiner_2017].  Sequencing of environmental DNA extracted from field-collected soil, water, or sediment samples can yield insight into a range of questions, from profiling the composition of ancient plant and animal communities [@Pedersen_2014], to motoring populations of rare or endangered species [@Balasingham_2017]. As the cost of eDNA metabarcoding declines and sample collection techniques become more streamlined (e.g. @Thomas_2018), professional research scientists are increasingly using eDNA metabarcoding as a platform to engage a diversity of community partners, including natural resource managers, undergraduate students, and citizen scientists in primary research. However, developing robust and impactful community science programs that engage community partners in all steps of the research process remains a challenge.

eDNA metabarcoding-based projects work well for programs that partner researchers with community scientists because non-experts can be quickly trained to collect samples in the field, and because eDNA metabarcoding is an exciting framework for research pertinent to disciplines such as medicine, agriculture, ecology, and geography [@Deiner_2017]. Community partners in such programs can have heterogeneous backgrounds, ranging from curious members of the public for whom collecting samples in the field is their first scientific research experience (e.g. University of California’s CALeDNA program, http://www.ucedna.com/), to professional natural resource managers who regularly collaborate with research scientists (e.g. Center for Ocean Solutions’ eDNA project, https://oceansolutions.stanford.edu/project-environmental-dna). A key ingredient to promote sustained engagement of community partners in these programs is that they should be able to participate across multiple stages of the research project, not only in sample collection [@Pandya_2012; @ECSA_2015]. This can be a challenge for eDNA metabacoding-based community science programs because although it is relatively easy to train community partners to collect eDNA samples, it is far more challenging to train them to independently visualize and analyze results from these studies. Indeed, learning the bioinformatic tools necessary for managing the large, multidimensional datasets generated in these studies can be difficult for professional researchers (Carey & Papin 2018), let alone for the non-technical audience of some community science programs.

To address this challenge, we created an R package “ranacapa”, at the core of which is a Shiny webapp that can be used to visualize results from eDNA sequencing studies and perform simple community ecology analyses. `ranacapa` complements existing visualization platforms (e.g. Phinch, Phyloseq-Shiny, QIIME2 Viewer), because in addition to interactive visualizations, `ranacapa` includes brief explanations of several core analyses used in eDNA studies and includes links to additional educational resources. `ranacapa` works with community matrices generated via QIIME [@Caporaso_2010] or the Anacapa sequence analysis pipeline (https://github.com/limey-bean/Anacapa), the latter being used extensively by the CALeDNA program.

@@

## Implementation
At the core of `ranacapa` is a Shiny webapp (@Chang_2018), which is available at http://gauravsk.shinyapps.io/ranacapa or with `ranacapa::runRanacapa()`. The package also includes two categories of helper functions (Table 1) that transform user-uploaded taxonomy and metadata tables into `R` objects that can be visualized and analyzed using the Phyloseq [@McMurdie_2013] and Vegan [@Oksanen_2018] packages. `ranacapa` is available for installation from Github or CRAN:

@@

- **Taxonomy heatmap**: Shows the taxon-by-sample matrix as an interactive heatmap made using `heatmaply::heatmaply()`, where the color of each cell represents the number of times a given taxon was sequenced in a sample (Figure 2). Users can filter the taxon list by selecting or deselecting specific taxa.

@@

- **Beta diversity statistics**: Shows results from two statistical tests of species turnover across site. The first statistical test is a multivariate ANOVA implemented with `vegan::adonis())`, which shows results from a statistical test of taxon turnover across sites. The second statistical test, which is implemented with `vegan::betadisper()`, is of heterogeneity of variances among samples. This test compares the degree of sample-to-sample variation within habitats (or within other user-selected groups).

@@

```{r print-taxon-table-structure}
set.seed(123)
# randomly select 10 samples to show
samples <- runif(10, min = 1, max = nrow(taxDat))
taxDat[samples,1:6]
```

@@

```{r print-medata-table-structure}
metDat[which(metDat[,1] %in% colnames(taxDat)[2:6]),1:5]
```

@@

## Use Cases
We expect that researchers with bioinformatic expertise will use the sequence analysis pipeline of their choice to assign taxonomy to eDNA datasets, and generate clean taxonomy and metadata files that can be visualized in `ranacapa`. Researchers can share these files with their partners, and emphasize the analyses or visualizations most appropriate to their use case. We now show how `ranacapa` can facilitate authentic communication between researchers and community partners in two settings.

@@

A team of UCLA researchers partnered with resource managers at the Channel Islands National Park Service to assess the potential for eDNA as a biodiversity monitoring tool to supplement expensive and time-intensive visual biodiversity surveys in the Southern California Channel Islands [@Lessios_1996; @Usseglio_2015; @Deiner_2017]. For this partnership, resource managers collected and filtered thirty unique one-liter water samples for eDNA analysis at permanent monitoring sites inside and adjacent to protected areas, and research scientists at UCLA performed eDNA sequencing of the mitochondrial 12S [@Miya_2015] and CO1 [Leray_2013] genes, targeting bony fishes, elasmobranches, and invertebrate taxa. The researchers processed sequences and assigned taxonomy using the Anacapa pipeline, and shared results with the resource managers using the `ranacapa` Shiny app.

The taxonomy heatmap of species detected using the 12S and CO1 metabarcodes (Figure 2) was the most valuable visualization to this collaboration, because it allowed the resource managers to filter the large observed species list down to a particular set of key taxa that they regularly monitor. The heatmap showed that this pilot study detected 36 of the 70 key metazoans at the species level, and the remaining 34 at the genus, family, or order level. This indicates that eDNA-based studies can provide critical information for ongoing management efforts and provide new insights into the spatial and temporal distributions of these species. The value of `ranacapa` in this scenario was to quickly sort through long species lists generated in by eDNA sequencing to highlight the strengths and weaknesses in using eDNA to monitor diversity in the Channel Islands. The data from this study are packaged as the demo dataset for the `ranacapa` Shiny app and are available online at XXX.

@@

A team of community ecology and environmental DNA researchers in the CALeDNA program collaborate with instructors of a research-based environmental microbiology course at UCLA [@Shapiro_2015], in which students used eDNA metabarcoding to study the impact of a local wildfire on the plant and soil microbial community. The goal of this twenty-week course was to provide undergraduate students an authentic experience in basic microbiology and microbial community ecology research. Over the first ten weeks, eDNA researchers on the instructional team sequenced the ITS2 [@Gu_2013] and 16S SSU RNA [@Caporaso_2012] metabarcoding regions from student=collected soil samples and used the Anacapa pipeline to generate taxon-by-sample tables.

The course instructors used the `ranacapa` Shiny app to introduce students to the structure of eDNA sequencing results. The students were encouraged to explore data the statistical analyses most pertinent to the hypotheses they had formed at the beginning of the course. A key benefit of using `ranacapa` was that despite having no prior bioinformatics experience, students could begin exploring the biodiversity in their samples in a matter of minutes by using the online instance of Shiny app. This allowed the instructors to focus classroom time on biological questions rather than on troubleshooting bioinformatics problems, as had been the case in previous sessions of the course. The course instructors noted that visualizing eDNA data in `ranacapa` helped students understand the relationships between taxon-by-site matrices and the various metadata they had collected in the field. By significantly reducing the time and difficulty in visualizing basic biodiversity patterns, `ranacapa` helped students develop and pursue more sophisticated analyses during the remainder of the course, using most sophisticated tools such as STAMP [@Parks_2014] and PICRUSt [@Langille_2013]. The taxonomy tables and metadata files used in this course are available online at XXXX.

@@

- A Shiny app, including a dataset generated for demonstrations, is available at https://gauravsk.shinyapps.io/ranacapa  
- `ranacapa` is available for installation at https://github.com/gauravsk/ranacapa  
- Link to source code as at time of publication (F1000Research TO GENERATE)  
- Link to archived source code as at time of publication (F1000Research TO GENERATE)  
- Software license (GPL-3)  
- Data availability   
- Figshare: [DOI]  
- License: CC-BY 4.0

@@

## Grant information
GSK and ZJG were supported by the US-NSF Graduate Research Fellowship (DEG No. 1650604). EEC, RSM, and the CALeDNA program are supported by the University of California President’s Research Catalyst Award (CA-16-376437).

@@

#### Figure 1
```{r, fig.cap="Figure 1: Taxon accumulation curve as shown in the first tab of ranacapa. "}
fig1
```

#### Figure 2
```{r, fig.cap="Figure 2: Taxonomy heatmap as shown in the `ranacapa` Shiny app. Taxonomy is shown at the Order level in this figure; in the app, users can choose the taxonomic level to show in the heatmap. Users can also filter  select or deselect individual taxa to be shown in the heatmap."}
fig2
```
#### Figure 3
```{r, fig.cap="Figure 3: Taxonomy barplot as shown in the `ranacapa` Shiny app. Taxonomy is shown at the Order level in this figure; in the app, users can choose the taxonomic level to show in the barplot"}
fig3
```

#### Figure 4
```{r, fig.cap="Figure 4: Alpha diversity boxplots as shown in the `ranacapa` Shiny app. Users can select the X-axis variable using a dropdown menu in the app."}
fig4
```
#### Figure 5
```{r, fig.cap="Figure 5: PCoA ordination of the samples as shown in the `ranacapa` Shiny app. Users can select the grouping variable with a dropdown menu in the app."}
fig5
```

>>>>

---
title: 'ranacapa: An R package and Shiny web app to explore environmental DNA data with exploratory statistics and interactive visualizations'
author: Gaurav S. Kandlikar (1,+), Zachary J. Gold (1), Madeline C. Cowen (1), Rachel  S. Meyer (1), Amanda C. Freise (2), Nathan J.B. Kraft (1), Jordan Moberg-Parker (2), Joshua Sprague (3), David Kushner (3), and Emily E. Curd (1)
fontsize: 11pt
output: 
  word_document:
    reference_docx: rendering-templates/word_template.docx
urlcolor: blue
always_allow_html: yes
bibliography: bibliography.bib
csl: rendering-templates/ecology-letters.csl
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, eval = F)
load("make-figures.Rdata")
options(digits = 3)
```

@@

### Affiliations and Contact Information  
(1) Department of Ecology and Evolutionary Biology, University of California – Los Angeles   
(2) Department of Microbiology and Molecular Genetics, University of California – Los Angeles  
(3) Channel Islands National Park, National Park Service, Ventura, California

@@

**Keywords**: environmental DNA, data visualization, citizen science, community science, shiny, metabarcoding, education, community ecology

@@

## Abstract 
```{r child = 'abstract.md', eval = T}
```

@@

The targeted amplification and sequencing of DNA that living organisms shed into their physical environment, termed "environmental DNA (eDNA) metabarcoding",  is revolutionizing microbiology, ecology, and conservation research [@Taberlet_2012; @Deiner_2017].  Sequencing of eDNA extracted from field-collected soil, water, or sediment samples can yield insight into a range of questions, from profiling the composition of ancient plant and animal communities [@Pedersen_2014], to monitoring populations of rare or endangered species [@Balasingham_2017]. As the cost of eDNA metabarcoding declines and sample collection techniques become more streamlined (e.g. @Thomas_2018), professional research scientists are increasingly using eDNA metabarcoding as a platform to engage a diversity of community partners, including natural resource managers, undergraduate students, and citizen scientists in primary research. However, developing robust and impactful community science programs that engage community partners in all steps of the research process remains a challenge.

eDNA metabarcoding-based projects work well for programs that partner researchers with community scientists because non-experts can be quickly trained to collect samples in the field, and because eDNA metabarcoding is an exciting framework for research pertinent to disciplines such as medicine, agriculture, ecology, and geography [@Deiner_2017]. Community partners in such programs can have heterogeneous backgrounds, ranging from curious members of the public for whom collecting samples in the field is their first scientific research experience (e.g. University of California’s CALeDNA program, http://www.ucedna.com/), to professional natural resource managers who regularly collaborate with research scientists (e.g. Center for Ocean Solutions’ eDNA project, https://oceansolutions.stanford.edu/project-environmental-dna). A key ingredient to promote sustained success of such programs is that community partners should be able to engage across multiple stages of the research project, not only in sample collection [@Pandya_2012; @ECSA_2015]. This can be a challenge for community science programs because although it is relatively easy to train community partners to collect eDNA samples, it is far more challenging to train them to independently visualize and analyze results from these studies. Indeed, learning the bioinformatic tools necessary for managing the large, multidimensional datasets generated in these studies can be difficult for professional researchers [@Carey_2018], let alone for the non-technical audience of some community science programs.

To address this challenge, we created the R package “`ranacapa`”, at the core of which is a Shiny web app that can be used to visualize results from eDNA sequencing studies and perform simple community ecology analyses. `ranacapa` complements existing visualization platforms (e.g. Phinch, Phyloseq-Shiny, QIIME2 Viewer), because in addition to interactive visualizations, `ranacapa` includes brief explanations of several core analyses used in eDNA studies and includes links to additional educational resources. `ranacapa` works with community matrices generated via QIIME [@Caporaso_2010] or the Anacapa sequence analysis pipeline (https://github.com/limey-bean/Anacapa), the latter being used extensively by the CALeDNA program.

@@

## Implementation
At the core of `ranacapa` is a Shiny web app [@Chang_2018], which is available at http://gauravsk.shinyapps.io/ranacapa or with `ranacapa::runRanacapa()`. The package also includes two categories of helper functions (Table 1) that transform user-uploaded taxonomy and metadata tables into `R` objects that can be visualized and analyzed using the Phyloseq [@McMurdie_2013] and Vegan [@Oksanen_2018] packages. `ranacapa` is available for installation from Github or CRAN:

@@

- **Taxonomy heatmap**: Shows the taxon-by-sample matrix as an interactive heatmap made using `heatmaply::heatmaply()`  [@Galili_2017], where the color of each cell represents the number of times a given taxon was sequenced in a sample (Figure 2). Users can filter the taxon list by selecting or deselecting specific taxa.

@@

- **Beta diversity statistics**: Shows results from two statistical tests of species turnover across site. The first test is a multivariate ANOVA of taxon turnover across sites, implemented with `vegan::adonis())`. The second statistical test, which is implemented with `vegan::betadisper()`, is of heterogeneity of variances among samples. This test compares the degree of sample-to-sample variation within habitats (or within other user-selected groups).

@@

```{r print-taxon-table-structure, eval = T, comment=""}
set.seed(123)
# randomly select 10 samples to show
samples <- runif(10, min = 1, max = nrow(taxDat))
print(taxDat[samples,1:5], row.names = FALSE)
```

@@

```{r print-medata-table-structure, eval = T, comment = ""}
print(metDat[which(metDat[,1] %in% colnames(taxDat)[2:5]),1:5], row.names = FALSE)
```

@@

## Use Cases
We expect that researchers with expertise in bioinformatics will use the sequence analysis pipeline of their choice to assign taxonomy to eDNA datasets, and generate clean taxonomy and metadata files that can be visualized in `ranacapa`. Researchers can share these files with their partners, and emphasize the analyses or visualizations most appropriate to their use case. We now show how `ranacapa` can facilitate authentic communication between researchers and community partners in two settings.

@@

A team of UCLA researchers partnered with resource managers at the Channel Islands National Park Service to assess the potential for eDNA as a biodiversity monitoring tool to supplement time-intensive visual biodiversity surveys in the Southern California Channel Islands [@Lessios_1996; @Usseglio_2015; @Deiner_2017]. For this partnership, resource managers collected and filtered 30 unique one-liter water samples for eDNA analysis at permanent monitoring sites inside and adjacent to protected areas, and research scientists at UCLA performed eDNA sequencing of the mitochondrial 12S [@Miya_2015] and CO1 [@Leray_2013] genes, targeting bony fishes, elasmobranches, and invertebrate taxa. The researchers processed sequences and assigned taxonomy using the Anacapa pipeline, and shared results with the resource managers using the `ranacapa` Shiny app.

The taxonomy heatmap of species detected using the 12S and CO1 metabarcodes (Figure 2) was the most valuable visualization to this collaboration, because it allowed the resource managers to filter the large observed species list down to a particular set of key taxa that they regularly monitor. The heatmap showed that this pilot study detected 36 of the 70 key metazoans at the species level, and the remaining 34 at the genus, family, or order level. This indicates that eDNA-based studies can provide critical information for ongoing management efforts and provide new insights into the spatial and temporal distributions of these species. The value of `ranacapa` in this scenario was to quickly sort through long species lists generated in by eDNA sequencing to highlight the strengths and weaknesses in using eDNA to monitor diversity in the Channel Islands. The data from this study are packaged as the demo dataset for the `ranacapa` Shiny app and are available online [@Kandlikar_2018a].

@@

A team of community ecology and environmental DNA researchers in the CALeDNA program collaborated with instructors of a research-based environmental microbiology course at UCLA [@Shapiro_2015], in which students used eDNA metabarcoding to study the impact of a local wildfire on the plant and soil microbial community. The goal of this twenty-week course was to provide undergraduate students an authentic experience in basic microbiology and microbial community ecology research. Over the first ten weeks, eDNA researchers on the instructional team sequenced the ITS2 [@Gu_2013] and 16S SSU RNA [@Caporaso_2012] metabarcoding regions from student-collected soil samples and used the Anacapa pipeline to generate taxon-by-sample tables.

The course instructors used the `ranacapa` Shiny app to introduce students to the structure of eDNA sequencing results. The students were encouraged to explore data and perform the statistical analyses most pertinent to the hypotheses they had formed at the beginning of the course. A key benefit of using `ranacapa` was that despite having no prior bioinformatics experience, students could begin exploring the biodiversity in their samples in a matter of minutes by using the online instance of Shiny app. This allowed the instructors to focus classroom time on biological questions rather than on troubleshooting bioinformatics problems, as had been the case in previous sessions of the course. The course instructors noted that visualizing eDNA data in `ranacapa` helped students understand the relationships between taxon-by-site matrices and the various metadata they had collected in the field. By significantly reducing the time and difficulty in visualizing basic biodiversity patterns, `ranacapa` helped students develop and pursue more sophisticated analyses during the remainder of the course, using tools such as STAMP [@Parks_2014] and PICRUSt [@Langille_2013]. The taxonomy tables and metadata files used in this course are available online [@Kandlikar_2018b].

@@

- A Shiny app, including a dataset generated for demonstrations, is available at https://gauravsk.shinyapps.io/ranacapa  
- `ranacapa` is available for installation at https://github.com/gauravsk/ranacapa  
- Link to source code as at time of publication (F1000Research TO GENERATE)  
- Link to archived source code as at time of publication (F1000Research TO GENERATE)  
- Software license (GPL-3)  
- Data availability   
- Figshare: https://doi.org/10.6084/m9.figshare.7199477 and https://doi.org/10.6084/m9.figshare.7199510.v1  
- License: CC-BY 4.0

@@

## Grant information
GSK and ZJG were supported by the US-NSF Graduate Research Fellowship (DEG No. 1650604). NJBK was supported the National Science Foundation DEB-1644641. EEC, RSM, and the CALeDNA program are supported by the University of California President’s Research Catalyst Award (CA-16-376437).

@@

#### Figure 1
Taxon accumulation curve as shown in the first tab of ranacapa. 
```{r}
fig1
```

#### Figure 2
Taxonomy heatmap as shown in the `ranacapa` Shiny app. Taxonomy is shown at the Order level in this figure; in the app, users can choose the taxonomic level to show in the heatmap. Users can also select or deselect individual taxa to be shown in the heatmap.
```{r}
fig2
```

#### Figure 3
Taxonomy barplot as shown in the `ranacapa` Shiny app. Taxonomy is shown at the Order level in this figure; in the app, users can choose the taxonomic level to show in the barplot
```{r}
fig3
```

#### Figure 4
Alpha diversity boxplots as shown in the `ranacapa` Shiny app. Users can select the X-axis variable using a dropdown menu in the app.
```{r}
fig4
```

#### Figure 5
PCoA ordination of the samples as shown in the `ranacapa` Shiny app. Users can select the grouping variable with a dropdown menu in the app.

```{r}
fig5
```
